---
title: '**SPRT and Poisson Calculations**'
author: "Joe Shaw"
date: "29/09/2021"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Rearranging SPRT

The SPRT calculations commonly used in relative mutation dosage are sourced from 
the supplementary material of Lo et al 2007 (PMID: 17664418).
The calculations can be accessed here (section: "Construction of SPRT Curves"):
<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1934923/bin/pnas_0705765104_index.html>.

The equations given calculate whether the likelihood ratio of a result exceed that of a predefined boundary, which is often given as 8.

upper boundary = [(ln 8)/N - ln d]/ln g

To calculate the likelihood ratio (lr) for each result, I have rearranged the above equation. ln (the natural logarithm) is the same as log to the base e, and I have written it as log because this is how it appears in R.

overrep_fraction = ((log(lr)/total_copies) - log(delta))/log(gamma)

overrep_fraction * log(gamma) = (log(lr)/total_copies) - log(delta)

(overrep_fraction * log(gamma)) + log(delta) = log(lr)/total_copies

(overrep_fraction * log(gamma)) + log(delta) * total_copies = log(lr)

Then apply laws of logs:

ln(x) = y 

x = e^y

lr <- exp((((overrep_fraction * log(gamma)) + log(delta)) * total_copies))

``` {r calc_lr, echo = TRUE}

calc_q1_autosomal <- function(ff) {
  q1 <-  0.5+(ff/2)
  return(q1)
}

calc_delta <- function(q0, q1) {
  delta <- (1- q1)/(1-q0)
  return(delta)
}

calc_gamma <- function(q0, q1) {
  gamma <- ((q1 * (1-q0))/ (q0*(1-q1)))
  return(gamma)
}

# Equation from Lo et al (2007)
calc_lr_boundary <- function(ff, total_copies) {
  q0 = 0.5
  q1 <- calc_q1_autosomal(ff)
  delta <- calc_delta(q0, q1)
  gamma <- calc_gamma(q0, q1)
  boundary <-  (log(8)/total_copies - log(delta))/log(gamma)
  return(boundary)
}

# My rearranged version
calc_lr_autosomal <- function(ff, overrep_fraction, total_copies) {
  q0 = 0.5
  q1 <- calc_q1_autosomal(ff)
  delta <- calc_delta(q0, q1)
  gamma <- calc_gamma(q0, q1)
  lr <- exp((((overrep_fraction*log(gamma)) + log(delta))*total_copies))
  return(lr)
}

```

## Testing the Rearrangement

By testing the rearranged equation on random data, the likelihood ratio should be calculated as 8 for every case by my rearranged version.

```{r test_data1, echo=TRUE}
test_data1 <- data.frame(
  test_sample = seq(from = 1, to = 20, by = 1),
  ff = sample(seq(from = 0.01, to = 0.15, by= 0.001), 20),
  total_copies = sample(seq(from = 100, to = 15000, by= 1), 20)) %>%
  mutate(
    boundary = calc_lr_boundary(ff, total_copies),
    lr = calc_lr_autosomal(ff, boundary, total_copies))

test_data1
```


## Simplifying SPRT 

For an autosomal recessive condition, the q0 value is always 0.5. Consequently,
the calculations for the delta, gamma and likelihood ratio values can be
simplified.


```{r lr_functions, include=TRUE}
calc_lr <- function(fetal_fraction, overrep_fraction, total_copies) {
  q0 <- 0.5
  q1 <- 0.5+(fetal_fraction/2)
  delta <- (1- q1)/(1-q0)
  gamma <- ((q1 * (1-q0))/ (q0*(1-q1)))
  lr <- exp((((overrep_fraction*log(gamma)) + log(delta))*total_copies))
  return(lr)
}

calc_lr_new <- function(fetal_fraction, overrep_fraction, total_copies) {
  delta <- 1 - fetal_fraction
  gamma <- (1 + fetal_fraction) / (1 - fetal_fraction)
  lr <- (gamma^overrep_fraction * delta)^total_copies
  return(lr)
}

```

## Testing the Simplification

By creating a random data-set of sensible input values, we can test if the
simplified function gives the same output as the original function.

```{r test_data2, echo=TRUE}
test_data2 <- data.frame(
  test_sample = seq(from = 1, to = 20, by = 1),
  fetal_fraction = sample(seq(from = 0.01, to = 0.15, by= 0.001), 20),
  overrep_fraction = sample(seq(from = 0.4, to = 0.6, by= 0.001), 20),
  total_copies = sample(seq(from = 100, to = 15000, by= 1), 20)) %>%
  
  mutate(
    lr_old = calc_lr(fetal_fraction, overrep_fraction, total_copies),
    lr_new = calc_lr_new(fetal_fraction, overrep_fraction, total_copies),
    difference = abs(lr_old - lr_new)) %>%
  
  select(test_sample, lr_old, lr_new, difference)

test_data2
```

## Reversing the Poisson correction

The Poisson correction when applied to digital PCR allows the calculation of the number of molecules of a given target based on the number of positive partitions for that target (P) and the total number of partitions examined (N).

When testing the appropriateness of the MCMC pipeline on genomic DNA data, I wanted to create an artificial data-set of fetal fractions.

To convert from a desired number of molecules to an expected number of positive droplets, this would require rearranging the standard Poisson distribution calculation:

```{r poisson_correct, include = TRUE}

poisson_correct <- function(N, P) {
  
  num_molecules <- -log((N-P)/N)*N
  return(num_molecules)
  
}
```

This requires three laws of natural logarithms

1) ln(x/y) = ln(x) - ln(y)
2) −ln(x) = ln(1/x)
3) If y = ln(x), then e^y = x

Rearranging using laws 1 and 2:

-ln((N-P)/N) = ln(1/ ((N-P)/N)) = ln(N/(N-P))

num_molecules  = ln(N/(N-P)) * N

num_molecules / N = ln(N/(N-P))

Rearranging using law 3 gives the number of predicted positive partitions:

e ^ (num_molecules / N) = N/(N-P)

N / (e ^ (num_molecules / N)) = N-P

N - N/ (e ^ (num_molecules / N)) = P

```{r reverse_poisson, include=TRUE}

reverse_poisson <- function(num_molecules, N) {
  
  P <- N - (N / exp(num_molecules/N))
  
  return(P)
}

```

## Testing the Rearrangement

Testing the rearranged equation on a randomly generated data-set.

```{r test_data3, echo=TRUE}

test_data3 <- data.frame(
  test_sample = seq(from = 1, to = 20, by = 1),
  # Total droplets
  N = sample(seq(from = 15000, to = 22000, by= 100), 20),
  # Positive droplets
  P = sample(seq(from = 10, to = 5000, by= 10), 20)) %>%
  
  mutate(
    num_molecules = poisson_correct(N, P),
    P_reverse = reverse_poisson(num_molecules, N),
    difference = abs(P - P_reverse))

test_data3
```

